---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
setwd('~/src/RepData_PeerAssessment1')
rawData = read.csv(file='activity.csv')
# remove NA
d = rawData[complete.cases(rawData), ]
```


## What is mean total number of steps taken per day?
```{r}
stepsByDay = aggregate(d$steps, by=list(d$date), FUN=sum)
hist(stepsByDay$x, xlab='Steps per day', main='Histogram of steps per day')
meanStepsPerDay = mean(stepsByDay$x)
medianStepsPerDay = median(stepsByDay$x)
```
- The mean total number of steps per day is `r sprintf("%.2f", meanStepsPerDay)`.
- The median number of steps per day is `r medianStepsPerDay`.


## What is the average daily activity pattern?
```{r}
stepsByTimeOfDay = aggregate(d$steps, by=list(d$interval), FUN=mean)
plot(x = stepsByTimeOfDay$Group.1, y = stepsByTimeOfDay$x, type = "l", 
     xlab = "Time of day", ylab = "Mean number of steps")
maxInterval = stepsByTimeOfDay[which.max(stepsByTimeOfDay$x),1]
```
- The maximum number of steps is contained in interval `r sprintf("%04d - %04d", maxInterval, maxInterval + 5)`

## Imputing missing values

```{r}
numMissingValues = sum(is.na(rawData$steps))
```
- Number of missing values = `r numMissingValues`

- Strategy for filling in NA data: take from mean steps by time of day:
```{r}
dWithImputed = rawData
dWithImputed[is.na(dWithImputed$steps),1] = stepsByTimeOfDay$x[match(dWithImputed[is.na(dWithImputed$steps),3], stepsByTimeOfDay$Group.1)]
stepsByDayWithImputed = aggregate(dWithImputed$steps, by=list(dWithImputed$date), FUN=sum)
hist(stepsByDayWithImputed$x, xlab='Steps per day', main='Histogram of steps per day, with imputed values')
meanStepsPerDayWithImputed = mean(stepsByDayWithImputed$x)
medianStepsPerDayWithImputed = median(stepsByDayWithImputed$x)
```
- The mean total number of steps per day is `r sprintf("%.2f", meanStepsPerDayWithImputed)`.
- The median number of steps per day is `r sprintf("%.2f", medianStepsPerDayWithImputed)`.

Obviously, as the imputed values were taken from the mean, the mean value has not changed. The median value has become equal to the mean.


## Are there differences in activity patterns between weekdays and weekends?
```{r, fig.height=8}
weekDayEndFactor = factor(ifelse(weekdays(as.Date(dWithImputed$date)) %in% c("Saturday", "Sunday"), "weekend", "weekdays"))
stepsByTimeOfDayWithImputed = aggregate(dWithImputed$steps, by=list(dWithImputed$interval, weekDayEndFactor), FUN=mean)

layout(matrix(c(1,2)), heights = c(2,2))
with (stepsByTimeOfDayWithImputed[stepsByTimeOfDayWithImputed$Group.2 == "weekend", ],
  plot(x = Group.1, y = x, type = "l", xlab = "Interval", ylab = "Number of steps", main="weekend")
)

with (stepsByTimeOfDayWithImputed[stepsByTimeOfDayWithImputed$Group.2 == "weekdays", ],
  plot(x = Group.1, y = x, type = "l", xlab = "Interval", ylab = "Number of steps", main="weekday")
)
```

The weekend number of steps are lower, more spread throughout the day and the morning peak is a bit later.